Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-23 manuscript/draft.Rmd
refs w/o conflicts

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
[SDMs; @thomas2004; @urban2015]. In general, these models describe individual
species' occurrence patterns as functions of the environment. Predictions about
where each species will occur in the future are made by using forecasts of
future environmental conditions to generate predictions from these models. These
species-level predictions are then combined ("stacked") to generate forecasts
for species richness [e.g. @calabrese2014]. Alternatively, models that directly
relate spatial patterns of species richness to environment conditions have been
developed and generally perform equivalently to stacked SDMs [@algar2009;
@distler2015]. This approach is sometimes referred to as "macroecological"
modeling, because it models the larger-scale pattern (richness) directly
[@distler2015].

Despite the emerging interest in forecasting species richness and other aspects
of biodiversity [[refs]], little is known about how effectively we can
anticipate these dynamics. This is due in part to the long time scales over
which many ecological forecasts are applied [and the resulting difficulty in
assessing whether the predicted changes occurred; @dietze2016]. What we do know
comes from a small number of hindcasting studies, where models are built using
data on species occurrence and richness from the past and evaluated on their
ability to predict contemporary patterns [e.g., @algar2009; @distler2015]. These
studies are a valuable first step, but lack several components that are
important for developing forecasting models with high predictive accuracy, and
for understanding how well different methods can predict the future. These "best
practices" for effective forecasting and evaluation (Box 1) broadly involve: 1)
expanding the use of data to include biological and environmental time-series
[@tredennick2016]; 2) accounting for uncertainty in observations, processes,
model choice, and forecast evaluation [@clark2001; @dietze2017]; and 3)
conducting meaningful evaluations of the forecasts by hindcasting, archiving
short-term forecasts, and comparing forecasts to baselines to determine whether
the forecasts are more accurate than assuming the system is basically static
[@perretti2013].

@@

**Past environmental data.** Environmental data included a combination of
elevation, bioclimatic variables and a remotely sensed vegetation index (the
normalized difference vegetation index; NDVI), all of which are known to
influence richness and distribution in the BBS data [[ref]]. For each year in
the dataset, we used the 4 km resolution PRISM data [@prism] to calculate eight
bioclimatic variables identified as relevant to bird distributions
[@harris2015]: mean diurnal range, isothermality, max temperature of the warmest
month, mean temperature of the wettest quarter, mean temperature of the driest
quarter, precipitation seasonality, precipitation of the wettest quarter, and
precipitation of the warmest quarter. Satellite-derived NDVI, a primary
correlate of richness in BBS data [@hurlbert2002], was obtained from the NDIV3g
dataset with an 8 km resolution [@pinzon2014] and was available from 1981-2013.
Average summer (May, June, July) and winter (December, January, Feburary) NDVI
values were used as predictors. Elevation was from the SRTM 90m elevation
dataset [@jarvis2008] obtained using the R package raster
[@hijmans2016]. Because BBS routes are 40-km transects rather than point counts,
we used the average value of each environmental variable within a 40 km radius
of each BBS route's starting point.

@@

**Joint Species Distribution Model: mistnet.** Joint species distribution models
(JSDMs) are a new approach that makes predictions about the full composition of
a community instead of modeling each species independently as above
[@warton2015]. JSDMs remove the assumed independence among species and
explicitly account for the possibility that a site will be much more (or less)
suitable for birds in general (or particular groups of birds) than one would
expect based on the available environmental measurements alone. As a result,
JSDMs do a better job of representing uncertainty about richness than stacked
SDMs [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.

@@

These results raise a number of important issues. First, we should be skeptical
of predictions about future biodiversity from models that have not been
validated in terms of repeated observations at the same site. Since site-level
richness is relatively stable [[refs]], model evaluations that show high
accuracy across spatial gradients may not indicate much about a model's ability
to predict changes over time. This issue has also recently been highlighted as a
potential downside of using SDMs to predict the future locations of individual
species [@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering
because SDMs form the main foundation for estimates of the predicted loss of
biodiversity to climate change [@thomas2004; @urban2015].

@@

One of the main alternatives to stacked SDMs is modeling richness directly using
environmental variables and using the resulting models to make forecasts. Two
studied compared the SDM and direct richness modeling approaches, and reported
that the methods yielded equivalent results for forecasting diversity
[@algar2009; @distler2015]. While our results generally support the idea of
rough equivalence between stacked SDMs and direct richness modeling for point
estimates, they also show that stacked SDMs dramatically understimate the
uncertainty. For this reason, direct modeling of richness is a better overall
approach to forecasting richness. A similar result is seen when comparing joint
species distribution models (JSDMs) to stacked single species distribution
models. For point estimates, the joint distribution models are roughly
equivalent to stacked SDMs, but the JSDMs provide much better estimates of
uncertainty [@harris2015; others?]. In fact, JSDMs and direct richness modeling
provide some of the best estimates of uncertainty across all modeling
approaches. This highlights the importance of evaluating models based on their
uncertainty as well as their mean estimates (Box 1). Not doing so is
particularly problematic in this case, because the summed species distribution
models are overly confident, which means that these models could mislead
ecologists into believing that richness would be restricted to a much narrower
range than would actually occur.

@@

Typically ecological forecasts use one modeling approach or a small number of
related approaches. By fitting and evaluating multiple modeling approaches we
can learn more rapidly about the best approaches for making predictions for a
given ecological quantity. This includes comparing process-based [e.g. [[refs]]]
and data-driven models, as well as comparing the accuracy of forecasts to simple
baselines to determine if the modeled forecasts are more accurate than the naive
assumption that the world is static [@ye2015].

@@

Many ecological forecasts use data that is readily available and easy to work
with. While ease of use is a reasonable consideration it is also important to
include predictor variables that are expected to relate to the ecological
quantity being forecast and dynamic time-series of predictors instead of
long-term averages. Investing time in identifying and acquiring better predictor
variables may have at least as many benefits as using more sophisticated
modeling techniques. [[Need a citation or two here]]

@@

Ecological observations are influenced by both the underlying biological
processes (e.g. resource limitation) and how the system is sampled. When
possible, forecasts should model the factors influencing the observation of the
data. [[citation?]]

>>>>

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
[SDMs; @thomas2004; @thuiller2011; @urban2015]. In general, these models
describe individual species' occurrence patterns as functions of the
environment. Predictions about where each species will occur in the future are
made by using forecasts of future environmental conditions to generate
predictions from these models. These species-level predictions are then combined
("stacked") to generate forecasts for species richness [e.g. @calabrese2014].
Alternatively, models that directly relate spatial patterns of species richness
to environment conditions have been developed and generally perform equivalently
to stacked SDMs [@algar2009; @distler2015]. This approach is sometimes referred
to as "macroecological" modeling, because it models the larger-scale pattern
(richness) directly [@distler2015].

Despite the emerging interest in forecasting species richness and other aspects
of biodiversity [@jetz2007; @thuiller2011], little is known about how
effectively we can anticipate these dynamics. This is due in part to the long
time scales over which many ecological forecasts are applied [and the resulting
difficulty in assessing whether the predicted changes occurred; @dietze2016].
What we do know comes from a small number of hindcasting studies, where models
are built using data on species occurrence and richness from the past and
evaluated on their ability to predict contemporary patterns [e.g., @algar2009;
@distler2015]. These studies are a valuable first step, but lack several
components that are important for developing forecasting models with high
predictive accuracy, and for understanding how well different methods can
predict the future. These "best practices" for effective forecasting and
evaluation (Box 1) broadly involve: 1) expanding the use of data to include
biological and environmental time-series [@tredennick2016]; 2) accounting for
uncertainty in observations, processes, model choice, and forecast evaluation
[@clark2001; @dietze2017]; and 3) conducting meaningful evaluations of the
forecasts by hindcasting, archiving short-term forecasts, and comparing
forecasts to baselines to determine whether the forecasts are more accurate than
assuming the system is basically static [@perretti2013].

@@

**Past environmental data.** Environmental data included a combination of
elevation, bioclimatic variables and a remotely sensed vegetation index (the
normalized difference vegetation index; NDVI), all of which are known to
influence richness and distribution in the BBS data [@kent2014]. For each year
in the dataset, we used the 4 km resolution PRISM data [@prism] to calculate
eight bioclimatic variables identified as relevant to bird distributions
[@harris2015]: mean diurnal range, isothermality, max temperature of the warmest
month, mean temperature of the wettest quarter, mean temperature of the driest
quarter, precipitation seasonality, precipitation of the wettest quarter, and
precipitation of the warmest quarter. Satellite-derived NDVI, a primary
correlate of richness in BBS data [@hurlbert2002], was obtained from the NDIV3g
dataset with an 8 km resolution [@pinzon2014] and was available from 1981-2013.
Average summer (May, June, July) and winter (December, January, Feburary) NDVI
values were used as predictors. Elevation was from the SRTM 90m elevation
dataset [@jarvis2008] obtained using the R package raster [@hijmans2016].
Because BBS routes are 40-km transects rather than point counts, we used the
average value of each environmental variable within a 40 km radius of each BBS
route's starting point.

@@

**Joint Species Distribution Model: mistnet.** Joint species distribution models
(JSDMs) are a new approach that makes predictions about the full composition of
a community instead of modeling each species independently as above
[@warton2015]. JSDMs remove the assumed independence among species and
explicitly account for the possibility that a site will be much more (or less)
suitable for birds in general (or particular groups of birds) than one would
expect based on the available environmental measurements alone. As a result,
JSDMs do a better job of representing uncertainty about richness than stacked
SDMs [@harris2015; @warton2015]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.

@@

These results raise a number of important issues. First, we should be skeptical
of predictions about future biodiversity from models that have not been
validated in terms of repeated observations at the same site. Since site-level
richness is relatively stable [[refs]], model evaluations that show high
accuracy across spatial gradients may not indicate much about a model's ability
to predict changes over time. This issue has also recently been highlighted as a
potential downside of using SDMs to predict the future locations of individual
species [@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering
because SDMs form the main foundation for estimates of the predicted loss of
biodiversity to climate change [@thomas2004; @thuiller2011; @urban2015].

@@

One of the main alternatives to stacked SDMs is modeling richness directly using
environmental variables and using the resulting models to make forecasts. Two
studied compared the SDM and direct richness modeling approaches, and reported
that the methods yielded equivalent results for forecasting diversity
[@algar2009; @distler2015]. While our results generally support the idea of
rough equivalence between stacked SDMs and direct richness modeling for point
estimates, they also show that stacked SDMs dramatically understimate the
uncertainty. For this reason, direct modeling of richness is a better overall
approach to forecasting richness. A similar result is seen when comparing joint
species distribution models (JSDMs) to stacked single species distribution
models. For point estimates, the joint distribution models are roughly
equivalent to stacked SDMs, but the JSDMs provide much better estimates of
uncertainty [@harris2015; @warton2015]. In fact, JSDMs and direct richness modeling
provide some of the best estimates of uncertainty across all modeling
approaches. This highlights the importance of evaluating models based on their
uncertainty as well as their mean estimates (Box 1). Not doing so is
particularly problematic in this case, because the summed species distribution
models are overly confident, which means that these models could mislead
ecologists into believing that richness would be restricted to a much narrower
range than would actually occur.

@@

Typically ecological forecasts use one modeling approach or a small number of
related approaches. By fitting and evaluating multiple modeling approaches we
can learn more rapidly about the best approaches for making predictions for a
given ecological quantity. This includes comparing process-based [e.g.
@kearney2009] and data-driven models, as well as comparing the accuracy of
forecasts to simple baselines to determine if the modeled forecasts are more
accurate than the naive assumption that the world is static [@ye2015].

@@

Many ecological forecasts use data that is readily available and easy to work
with. While ease of use is a reasonable consideration it is also important to
include predictor variables that are expected to relate to the ecological
quantity being forecast and dynamic time-series of predictors instead of
long-term averages. Investing time in identifying and acquiring better predictor
variables may have at least as many benefits as using more sophisticated
modeling techniques [@kent2014].

@@

Ecological observations are influenced by both the underlying biological
processes (e.g. resource limitation) and how the system is sampled. When
possible, forecasts should model the factors influencing the observation of the
data. [@yu2010; @hutchinson2011; @schurr2012]
